%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigplan')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%

\documentclass[sigplan,screen,review]{acmart}


\usepackage{paper_alcides}
\usepackage{cleveref}
\usepackage{pgfplots}
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{positioning}
\usepgfplotslibrary{fillbetween}
\usepgflibrary{patterns}
\usetikzlibrary{patterns}
\usepackage{bussproofs}
\usepackage{listings}
\usepackage{bbding}

\usepackage[framemethod=tikz]{mdframed}

\definecolor{mycolor}{rgb}{0.122, 0.435, 0.698}

\newmdenv[innerlinewidth=0.5pt, roundcorner=4pt,linecolor=mycolor,innerleftmargin=6pt, innerrightmargin=6pt,innertopmargin=6pt,innerbottommargin=6pt]{prittybox}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}


%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}


%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title[Implementing Separation Logic using an SMT-backed Frame Rule]{Implementing Separation Logic \\ using an SMT-backed Frame Rule}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Kirill Golubev}
\email{gkigorevich@ciencias.ulisboa.pt}
\orcid{0009-0002-2709-5241}

\author{Alcides Fonseca}
\email{alcides@ciencias.ulisboa.pt}
\orcid{0000-0002-0879-4015}
\affiliation{%
  \institution{LASIGE, Faculdade de Ciências da Universidade de Lisboa}
  \city{Lisboa}
  \country{Portugal}
}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

Symbolic execution is a technique frequently used to reason about code. In symbolic execution, an analyzer keeps track of the program state using its logic representation and validates the correctness of state transitions. This verification is usually discharged to SMT solvers as queries to the logical state.

Separation Logic is frequently used to express and verify the properties of programs with pointers or references. However, most SMT solvers (like the popular Z3) do not support Separation Logic natively. Recently, the CVC5 SMT Solver has introduced partial support for separation logic, which has not yet been integrated into more high-level tools.

This work aims to address this gap, by providing a proof of concept for implementing the Frame Rule using SMT queries in the Symbolic Heap fragment of Separation Logic, as supported by CVC5. We conclude that this encoding can simplify the machinery dealing with separation logic, such as that present in Smallfoot or Verifast.

\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10011742</concept_id>
<concept_desc>Theory of computation~Separation logic</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10003790.10011742</concept_id>
<concept_desc>Theory of computation~Separation logic</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10002944.10011123.10011676</concept_id>
<concept_desc>General and reference~Verification</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Separation logic}
\ccsdesc[500]{Theory of computation~Separation logic}
\ccsdesc[300]{General and reference~Verification}


\received{20 February 2007}
\received[revised]{12 March 2009}
\received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle


\newcommand{\EM}[1]{\ensuremath{#1}}
\newcommand{\ssymbol}[1]{\EM{#1}}
\newcommand{\bnfdef}{\EM{\vcentcolon\vcentcolon=}}
\newcommand{\emphbf}[1]{\textbf{\emph{#1}}}
\newcommand{\spmid}{\EM{\ \mid \ }}
\newcommand{\vsample}[1]{\EM{\mathit{sample}(#1)}}
\newcommand{\vconst}{\EM{\mathsf{c}}}
\newcommand{\anyval}{\ssymbol{v}}
\newcommand{\anydist}{\ssymbol{d}}
\newcommand{\dnormal}[2]{\EM{\mathit{Normal}(#1,#2)}}
\newcommand{\duniform}[2]{\EM{\mathit{Uniform}(#1,#2)}}

\lstset{mathescape}

\section{Introduction} 

Formal verification is one of the few known ways to ensure that a computer program has the least amount of errors. The idea is to prove that a given program satisfies a specification provided in advance. The language of program specification is usually some logic that is fitted to describe program behavior. There are many ways to achieve this with varying resource requirements and automation degrees.

One technique employed to verify imperative programs is symbolic execution~\cite{berdine2005symbolic}. Usually, the engine for it is implemented separately for each tool~\cite{DBLP:conf/fmco/BerdineCO05,DBLP:conf/oopsla/DistefanoP08}. In the context of this work, the state of the symbolic executor can be represented as a logical statement about the program. SMT solvers are used as oracles to verify that each transition of the symbolic execution engine's state is correct. 

Most programs in imperative languages are written in terms of heap manipulation. Separation Logic~\cite{DBLP:journals/cacm/OHearn19} is an extension of Hoare Logic~\cite{DBLP:journals/cacm/Hoare69}, that is frequently used to reason about these types of programs. It introduces some new operations and constants, alongside a new inference rule called \emph{Frame Rule}. In general, Separation Logic is proved to be undecidable~\cite{DBLP:conf/lics/BrotherstonK10}, but there are decidable subsets.

In this work, we focus on the \emph{Symbolic Heap} fragment~\cite{DBLP:conf/fmco/BerdineCO05}, as it enables a significant degree of automation by being supported in the CVC5 SMT Solver~\cite{DBLP:conf/tacas/BarbosaBBKLMMMN22}. A simplified Symbolic Heap requires the logical context to be split into a pure boolean part and a pure spatial part, made of disjoint heaplets. See \Cref{fig:framerule} for details.

This simplified version significantly restricts the expressive power of Separation Logic but still permits the encoding of some interesting program properties. Thus, we chose it as a starting point, but CVC5 actually supports a larger fragment called GRASS~\cite{DBLP:conf/cav/PiskacWZ13}\footnote{Actually, GRASS is a fragment of first-order logic which hosts the decidable fragment of separation logic. We will further use GRASS as a name for the decidable separation logic fragment for convenience.}. In contrast, Z3\cite{DBLP:conf/tacas/MouraB08} does not support separation logic.~\footnote{But it was once prototyped: \url{https://github.com/Z3Prover/z3/issues/811}.} 

\begin{figure}
\begin{prittybox}
\begin{align*}
    &\texttt{x, y, $\dots$} \in Variables &&\text{variables}\\
    &\texttt{$\alpha$, $\beta$, $\dots$} \in Values &&\text{values}\\
    &\texttt{S := true | x $\mapsto \alpha$} &&\text{simple spatial formula}\\
    &\texttt{$\Sigma$ := emp | S * $\Sigma$} &&\text{spatial formula}\\
    &\Pi \land \Sigma &&\text{simplified symbolic heap}
\end{align*}

\begin{prooftree}
    \AxiomC{\texttt{\{pre\} code \{post\}}}
    \RightLabel{FR}
    \UnaryInfC{\texttt{\{pre * frame\} code \{post * frame\}}}
\end{prooftree}
\caption{Simplified symbolic heap fragment of separation logic}
\end{prittybox}

\label{fig:framerule}
\end{figure}


The goal of this work is to highlight the opportunity to shift some heavy lifting related to separation logic from a symbolic execution engine to an SMT solver. For this, we present an algorithm to encode the Frame Rule of Separation Logic (\Cref{fig:framerule}) through SMT Solver queries.

\section{Related work}

Smallfoot~\cite{DBLP:conf/fmco/BerdineCO05} was the first tool to implement verification with the symbolic heap fragment of separation logic. The rules related to separation logic are complex, and it makes sense to delegate the verification to a dedicated tool, such as an SMT solver. 

GRASShopper~\cite{piskac2014grasshopper} is a tool that implements the GRASS fragment of separation logic for analyzing C-like language. In contrast to our approach, the engine for analysis is implemented from scratch instead of relying on an SMT solver.

Verifast~\cite{jacobs2011verifast} consumes and produces chunks of the heap relying on SMT proofs of pointer equivalence, which should result in a multiplicative amount of SMT queries with respect to symbolic heap size and number of heap chunks in an assertion.

\section{Frame Rule}

\paragraph{Preliminaries} To prove that a formula with universally free variables holds, we query the SMT solver whether the negation of the query, using existentially quantified free variables, is unsatisfiable (\texttt{unsat}). We denote this verification as $\operatorname{isUNSAT}(\neg\mbox{query})$.

The frame rule application is treated here as if it is in the context of function application. It is general enough to apply the same procedure in other contexts. \Cref{fig:FRAPP} is an example from Liquid Java~\cite{gamboa2021user} to illustrate the context.

\begin{figure}
    \centering
    \begin{lstlisting}[language=Java]
//definition
@HeapPrecondition("x $\mapsto$ ()")  //pre
@HeapPostcondition("_ $\mapsto$ ()") //post
Object foo(Object x){ ... }

//use site
void main(){
    //H$_{in}$ = emp
    var a = new Object();// FR app
    //H$_{out}$ = a $\mapsto$ ()
    
    //H$_{in}$ = a $\mapsto$ ()
    var b = new Object();// FR app
    //H$_{out}$ = a $\mapsto$ () * b $\mapsto$ () 
    
    //H$_{in}$ = a $\mapsto$ () * b $\mapsto$ () 
    //H$_{in}$ = pre * frame
    var r = foo(a); // FR app
    //H$_{out}$ = r $\mapsto$ () * b $\mapsto$ ()
    //H$_{out}$ = post * frame
}
\end{lstlisting}
    \caption{Frame rule application in Liquid Java. The pure boolean context is discarded for readability}
    \label{fig:FRAPP}
\end{figure}



\paragraph{Algorithm} The algorithm to check the Frame Rule is split into two phases: 

\begin{itemize}
    \item Check if the current context satisfies the precondition
    \item Apply postcondition to the larger context
\end{itemize}


The first step focus on verifying whether the precondition is guaranteed by the context.
The context is composed of the pure boolean context - $\mbox{BCtx}$, 
defining pointer equivalence and the heap description - 
$\mbox{H}_{in}$. The goal is to prove that 
$\mbox{H}_{in}$, contains precondition modulo pointer equivalence.

\begin{align*}
\mbox{BCtx} \wedge H_{in} \Vdash_{SMT} \mbox{pre} & \, \iff & \\ & \operatorname{isUNSAT}(\neg(\mbox{BCtx} \wedge H \implies \mbox{pre} \mathbin{*} \mbox{true} ))
\end{align*}

%\begin{lstlisting}[mathescape]
%BCtx | H $\Vdash$ pre 
% $iff$ isUNSAT($\neg$(BCtx $\land$ H $\implies$ pre * true))
%\end{lstlisting}


The second step isolates the unchanged part of the heap, called $\mbox{frame}$, to be kept in the outgoing context, by discarding all heaplets that invalidate the precondition. In particular, for the incoming context $S_{in} = \text{BCtx} \wedge H_{in} = \text{BCtx} \wedge (\text{pre} * \text{frame})$, we will generate the outgoing context $S_{out} = \text{BCtx} \wedge H_{out} = \text{BCtx} \wedge (\text{post} * \text{frame})$. 

\begin{align*}
&H_{in} = h_1 \mathbin{*} \dots \mathbin{*} h_n\\
&frame = \* \{ h_i  \,|\, i \in 1\dots n,\, \mbox{BCtx} \wedge H_{in} \Vdash_{SMT} \mbox{pre} \mathbin{*} h_i  \mathbin{*} \text{true} \}\\
&H_{out} = \text{post} \mathbin{*} \prod_{h_j \in frame} h_j = \text{post} \mathbin{*} \text{frame}
\end{align*}

Where $\prod$ is used with respect to separating conjunction. From this heap reconstruction, we can build the complete outgoing context $S_{out}$. The heap reconstruction process can be used in other contexts, such as merging heaps after branching.

When compared to other SMT-based verification procedures, our approach brings an overhead of \texttt{O(size(H))} SMT queries.

\paragraph{Evaluation} We validated the feasibility of this approach by implementing it in the \href{https://catarinagamboa.github.io/liquidjava.html}{Liquid Java} compiler~\cite{gamboa2021user}, supporting function calls, conditional branching, and assignments. The performance degradation for synthetic benchmarks is around 30\% relative to the pure boolean version of Liquid Java. Our implementation is available at \url{https://github.com/CatarinaGamboa/liquidjava/pull/20}.

\todo{maybe mention benchmarking details?}
\alcides{What programs were run, how complex were they? how long did the total verification take?}


\section{Conclusions}

The big advantage of this approach is that the SMT solver algorithm for separation logic is decidable, in contrast to Viper\cite{DBLP:conf/vmcai/0001SS16}, which is a popular backend for implementing reasoning in some ownership logic. 

In contrast to GRASShopper\cite{piskac2014grasshopper}, which includes a decidable inference algorithm, our approach provides a much more lightweight path to incorporating it into verification tools. Another contribution of our work is the exploration of the potential of separation logic support within SMT Solvers.


The simplicity and decidability come with the cost of features that are possible to support. Viper is a much more mature and rich backend for the language, while the presented approach is capped by the capabilities of separation logic support in SMT solver. Said capabilities are defined by GRASS fragment of separation logic which looks like ``propositional'' separation logic. The feature that is particularly challenging to support is fractional permissions~\cite{DBLP:conf/sas/Boyland03}. 

While our validation was in the specific context of Liquid Java, it is general enough to be used in other projects relying on SMT solvers to verify symbolic execution steps. The primary benefit of this algorithm is simplicity and delegation of responsibility for separation logic handling to the SMT solver instead of a symbolic execution engine which is usually implemented separately for each tool. 


\bibliographystyle{acm}
\bibliography{main}

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.